---
output: 
  html_document:
    toc: TRUE
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: flatly
    highlight: tango
    df_print: paged
    css: style2.css
---

<div class="casefile-wrapper">

  <div class="blood-stamp">CONFIDENTIAL CASE FILE</div>
  
  <a href="index.html" class="casefile-button">📁 Return to the Front</a>

  <h1 class="casefile-title">About the Crime Scene</h1>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#Loading libraries needed for this section 
library(tidyverse)
library(janitor)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Loading the data 
cancer <- read_csv("data/cancer_data.csv") %>% 
  janitor::clean_names()
```

### Understanding the Crime Scene
The crime scene is where the evidence is found. In this case, the crime scene is the biological and clinical data collected from patients diagnosed with breast cancer. This data includes various features such as tumor size, lymph node involvement, hormone receptor status, and patient demographics. These features help us understand the nature of the crime committed by breast cancer and provide insights into its behavior

